smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-09-09 manuscript/manuscript.qmd
incorporated PAS edits and comments

[**Word count:**]{.underline} 1,983 words/2,000 words

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes. Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data. If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied. Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold. Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders. The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
:::

@@

In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far only been described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].

@@

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be specified (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.

@@

The smdi R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.

@@

`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset.")`

```{r}
#| echo: false
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3,
  ) %>% 
  smdi_style_gt()
```

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
```

@@

```{r}
#| label: fig-examples
#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
#| echo: false
#| warning: false
#| fig-dpi: 400
#| fig-width: 15
#| fig-height: 10

# a) smdi_vis()
panel_a <- smdi_vis(data = smdi_data, strata = "exposure")
panel_a <- panel_a + ggplot2::theme_bw(base_size = 15)

@@

panel_b <- as.ggplot(smdi::gg_miss_upset(data = data_monotone))

#smdi_asmd
asmd <- smdi_asmd(data = smdi_data)
panel_c <- asmd$pdl1_num$asmd_plot

#smdi_asmd
rf <- smdi_rf(data = smdi_data, n_cores = 3)
panel_d <- rf$ecog_cat$rf_plot

panel <- cowplot::plot_grid(
  panel_a, panel_b, panel_c, panel_d,
  ncol = 2,
  nrow = 2,
  labels = c("a)", # % missing
             "b)", # set visualization
             "c)", # asmd
             "d)" # rf
             ),
  label_size = 18,
  label_colour = "black",
  label_fontface = "bold",
  vjust = 1.25,
  hjust = -.25
  )

@@

jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 16, height = 10, units = "in", res=400)
panel
invisible(dev.off())
```

@@

```{r}
#| label: fig-results
#| eval: true
#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
#| echo: false
knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
```
:::

>>>>

[**Word count:**]{.underline} 1,997 words/2,000 words

@@

Contrary to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes. Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data. If the underlying missingness mechanisms of such factors are not observed, for example, if patients with higher levels of a prognostic biomarker are more likely to have missing data for that biomarker, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied. Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold. Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders. The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
:::

@@

In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far only been described and tested in isolation from each other and no unified principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].

@@

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio, number of trees and CPU cores to parallelize over can be specified (`train_test_ratio, ntree` and `n_cores` parameter, respectively). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the model. In case of inflated AUC values (\>0.9), the function prompts a message to the user reporting the most important predictor. If in such a scenario another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to run the diagnostics for each POC independently rather than jointly.

@@

The `smdi` R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.

@@

`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset. In this example, ECOG performance score (ecog_cat) shows no imbalances in patient characteristics between patient with and without an observed value (absolute standardized mean difference [ASMD] 0.029, p[Hotelling] 0.783, group 1 diagnostic). Additionally missingness cannot be predicted well (AUC = 0. 510, group 2 diagnostic) and no difference in the outcome can be observed between patients with and without ecog_cat (log HR -0.06 [95% CI -0.16; 0.03], group 3 diagnostic). Accordingly, the missingness diagnostics indicate that ECOG follows a missing completely at random missingness (MCAR) mechanism. Similarly, the EGFR (egfr_cat) and PD-L1 (pdl1_num) biomarker variables can be interpreted as following a missing at random (MAR) and missing not at random value (MNARvalue) mechanism. See also @fig-results.")`

smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3
  ) %>% 
  smdi_style_gt()
```

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%

knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
```

@@

```{r}
#| label: fig-examples
#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively. Sub-figure a) displays the proportion of missing observations for each partially observed covariate stratified by exposure. The upset plot in sub-figure b) demonstrates how a monotone missingness pattern between partially observed covariates can be visually inspected using a set visualization technique [@ruddle2022]. Sub-figure c) illustrates absolute standardized mean differences (ASMD) in patient characteristics between patients with and without a value observed for the PD-L1 (pdl1_num) biomarker as a measure of imbalance. Sub-figure d) demonstrates the variable importance of fully observed covariates for predicting missingness in the partially oberserved ECOG performance score variable (ecog_cat)."
#| echo: false
#| warning: false
#| fig-dpi: 400
#| fig-width: 22
#| fig-height: 11

# a) smdi_vis()
panel_a <- smdi_vis(data = smdi_data, strata = "exposure")
panel_a <- panel_a + ggplot2::theme_bw(base_size = 18)

@@

panel_b <- as.ggplot(smdi::gg_miss_upset(
  data = data_monotone,
  point.size = 3.2, 
  text.scale = 2
  )
  ) +
  labs(
    title = "Upset plot iIllustrating missingness patterns among partially observed covariates"
  ) +
  theme(plot.title = element_text(hjust = 0.6, size = 20))

#smdi_asmd
asmd <- smdi_asmd(data = smdi_data)
panel_c <- asmd$pdl1_num$asmd_plot + 
  ggplot2::theme_bw(base_size = 18)

#smdi_asmd
rf <- smdi_rf(data = smdi_data, n_cores = 3)
panel_d <- rf$ecog_cat$rf_plot +
  ggplot2::theme_bw(base_size = 18)

panel <- cowplot::plot_grid(
  panel_a, panel_b, panel_c, panel_d,
  ncol = 2,
  nrow = 2,
  labels = c("a)", # % missing
             "b)", # set visualization
             "c)", # asmd
             "d)" # rf
             ),
  label_size = 20,
  label_colour = "black",
  label_fontface = "bold",
  vjust = 1.25,
  hjust = -.25
  )

@@

jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 22, height = 11, units = "in", res=400)
panel
invisible(dev.off())
```

@@

```{r}
#| label: fig-results
#| eval: true
#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
#| echo: false
#| fig-dpi: 400
#| fig-width: 22
#| fig-height: 11
knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
```
:::
